{\color{red} \sout{The Importance of Cell Connectivity for Efficient and }}Accurate {\color{red} \sout{Modeling} Simulation} of Flow through Dipping Aquifers with MODFLOW 6 {\color{red} Using Enhanced Cell Connectivity}

### 2024-11-19 latex/saved drafts/article.tex
intermediate review drafts & final draft in progress
intermediate review drafts & final draft in progress

With full connectivity and the standard conductance-based flow formulation (Figure \ref{fig:fig4}c), the magnitude of the error in the flux magnitude is less than 17\% of the analytical solution of 1 m/d, and the magnitude of the error in the flux direction is less than 10$^{\circ}$, at dip angles less than 40$^{\circ}$ for all of the conductivity ratios simulated. Given that the cells within the aquifer are square, cells that are nominally in the same grid layer within the aquifer no longer share any interfacial area, i.e., the become hydraulically disconnected, at a dip angle of 45$^{\circ}$. A second transition in connectivity occurs at a dip angle of approximately 63$^{\circ}$, and at a dip angle of approximately 72$^{\circ}$ the vertical columns of cells within the aquifer are completely disconnected from their neighbors to either side. When the discretization used to represent the aquifer becomes substantially disconnected, it cannot be expected to support accurate results. Note that for coarser horizontal discretization, transitions in connectivity within the aquifer would occur at smaller dip angles.

@@

\section{Supporting Information}
During review, the files that support the creation and running of the test problem are available at {\color{red} XXXXXXXXXX}. Once these files are finalized prior to publication, they will be made available at a permanent URL that will be noted here.

>>>>

With full connectivity and the standard conductance-based flow formulation (Figure \ref{fig:fig4}c), the magnitude of the error in the flux magnitude is less than 17\% of the analytical solution of 1 m/d, and the magnitude of the error in the flux direction is less than 10$^{\circ}$, at dip angles less than 40$^{\circ}$ for all of the conductivity ratios simulated. Given that the cells within the aquifer are square, cells that are nominally in the same grid layer within the aquifer no longer share any interfacial area, i.e., they become hydraulically disconnected, at a dip angle of 45$^{\circ}$. A second transition in connectivity occurs at a dip angle of approximately 63$^{\circ}$, and at a dip angle of approximately 72$^{\circ}$ the vertical columns of cells within the aquifer are completely disconnected from their neighbors to either side. When the discretization used to represent the aquifer becomes substantially disconnected, it cannot be expected to support accurate results. Note that for coarser horizontal discretization, transitions in connectivity within the aquifer would occur at smaller dip angles.

@@

\section{Supporting Information}
During review, the files that support the creation and running of the test problem are available at https://github.com/langevin-usgs/xt3d-vertically-staggered/releases/download/v1.0/suppinfo.zip. Once these files are finalized prior to publication, they will be made available at a permanent URL that will be noted here.

### 2024-11-19 latex/saved drafts/article_clean.tex
intermediate review drafts & final draft in progress
intermediate review drafts & final draft in progress

In simulations of groundwater flow through dipping aquifers, layers of model cells are often ``deformed'' to follow the top and bottom elevations of the aquifers. When this approach is used in MODFLOW, adjacent cells within the same model layer are vertically offset from one another, and the standard conductance-based (two-point) formulation for flow between cells does not rigorously account for these offsets. The XT3D multi-point flow formulation in MODFLOW 6 is designed to account for geometric irregularities in the grid, including vertical offsets, and to provide accurate results for both isotropic and anisotropic groundwater flow.  A recent study evaluated the performance of the standard formulation and XT3D using a simple, synthetic benchmark model of a steeply dipping aquifer. Although XT3D generally improved the accuracy of flow simulations relative to the standard formulation as expected, neither formulation produced accurate flows in cases that involved large vertical offsets. In this paper, we explain that the inability of XT3D to produce accurate flows in the steeply dipping aquifer benchmark was not due to an inherent limitation of the flow formulation, but rather to the limited cell connectivity inherent in the most commonly used discretization packages in MODFLOW 6. Furthermore, we demonstrate that XT3D is able to produce the expected accuracy when adequate cell connectivity is introduced using MODFLOW's unstructured grid type and the aquifer is discretized vertically using at least two model layers.

@@

Numerical simulation of layered aquifer systems can involve a tradeoff between efficiency and accuracy. A traditional approach to discretizing layered aquifer systems, called the ``grid overlay'' method by \cite{hoaglund2003}, maps the heterogeneous hydrologic properties onto a rectilinear grid \citep{modflow84}. Although this method preserves the benefits of the rectilinear grid structure, it can require a large number of model grid layers to adequately discretize hydrogeologic layers of variable dip and thickness \citep{Zyvoloski2006}.

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This commonly used method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. However, the non-orthogonality of the resulting grid can be a source of substantial model error \citep{hoaglund2003}.

@@

The introduction of unstructured grids into MODFLOW \citep{modflowusg, modflow6gwf} presented new ways for a grid to violate the CVFD requirements. To help overcome the limitations of the two-point flow formulation on geometrically irregular grids, MODFLOW-USG and MODFLOW 6 offer a Ghost-Node Correction (GNC) Package \citep{modflowusg, modflow6gwf}. However, configuration of optimal ghost node locations and interpolation weights is problem-dependent, and we are not aware of guidelines for handling general unstructured grids or arbitrarily oriented 3D anisotropy.

The optional XT3D flow formulation \citep{modflow6xt3d} in MODFLOW 6 offers a more ``automatic" alternative to ghost nodes. XT3D is a multi-point flow formulation \citep{edwards1998, aavatsmark2002} based on gradient reconstruction \citep{mavriplis2003leastsq, diskin2008accuracy}. By performing flow calculations using Darcy's Law in its tensorial form using a multi-point approximation of the head gradient vector, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid. In theory, XT3D should be able to compensate for the geometric irregularity introduced by vertical offsets between adjacent cells in the same grid layer.

@@

Versions of MODFLOW prior to MODFLOW-USG \citep{modflowusg} supported only structured grids composed of rows, columns, and layers of cells, which could be vertically offset. This grid type, called ``DIS," continues to be supported in MODFLOW-USG and MODFLOW 6, which also offer an unstructured grid type called ``DISU" \citep{modflowusg, modflow6gwf}. In addition, MODFLOW 6 offers an  unstructured grid type called ``DISV," which is based on specification of cell vertices \citep{modflow6gwf}. The DIS and DISV grid types are based on model layers, with the same horizontal grid structure within each layer. In a DIS or DISV grid, a connection between two cells in different model layers is created automatically if and only if one cell is directly below the other and neither cell has been optionally excluded from the model domain by the user. The DISU grid type is more flexible and, in MODFLOW 6, does not require the grid to have a layered structure. (The DISU grid type in MODFLOW-USG is similarly flexible but retains the concept of layers.) In a DISU grid, the user explicitly specifies all connections between grid cells. Thus, when used to represent a grid with a layered structure, as in this work, the DISU grid type allows connections to exist between any two cells, whether they are in the same layer or different layers. Grid geometry and connectivity information supplied by the discretization package (DIS, DISV, or DISU) is used by the MODFLOW 6 Node Property Flow (NPF) Package to formulate expressions for groundwater flow between cells in terms of heads calculated at cell centers. By default, the NPF Package uses a two-point flow formulation, which we call the ``conductance-based'' formulation. The optional, multi-point XT3D flow formulation can be activated simply by including the corresponding keyword in the NPF Package input.

@@

Vertically offset grids violate the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same grid layer are not strictly horizontal and therefore not perpendicular to the vertical cell-cell interfaces they intersect. For example, in Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell-cell interface, which reduces the accuracy of the conductance-based formulation.

@@

Specific discharge at cell centers is calculated in a postprocessing step using distance-weighted interpolation of face flows calculated by MODFLOW 6 and reported in the binary budget output file. For the purpose of illustrating and evaluating the results of this idealized benchmark, specific discharge in cells within the aquifer is calculated using only face flows between cells within the aquifer. Likewise, specific discharge in cells within the surrounding domain is calculated using only face flows between cells within the domain. These specific discharge values are different from those reported in the binary budget output file by MODFLOW 6 because MODFLOW interpolates specific discharge in a cell using all of the face flows for that cell, including flows across cell faces on the aquifer-domain boundary. The flows reported by MODFLOW represent an ``average'' of the different flows on either side of the aquifer-domain boundary. Therefore, specific discharge calculated in a cell adjacent to the aquifer-domain boundary is affected by the different flow rate on the other side of the boundary. Although this is an unavoidable consequence of vertically offset discretization in practical applications, and not necessarily a problem, users should be aware of these effects when interpreting the results of MODFLOW 6 dispersive-transport and MODPATH particle-tracking simulations.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows substantially improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c), which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively. Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d), reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.

@@

The results in Figure \ref{fig:fig3} confirm that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively. When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no substantial benefit in our benchmark problem (Figure \ref{fig:fig3}b).

@@

The base case and variations on the base case up to this point have used isotropic hydraulic conductivity. However, sedimentary layers often exhibit anisotropic conductivity. Therefore, we examine the effect of anisotropy within the aquifer and domain by dividing the original conductivity perpendicular to the dip of the aquifer by factors of 1, 10, 100, 1,000 and 10,000 to give anisotropy ratios ranging from 1.0 (isotropic) to 0.0001. Orientation of the principal directions of anisotropy parallel and perpendicular to the dip of the aquifer is accomplished by setting three angles in the NPF Package input. The first angle is the rotation within the horizontal plane of the first principal direction (typically the direction of maximum hydraulic conductivity) with respect to the positive $x$ axis. The second angle is the rotation of the first principal direction upward from the horizontal plane. The third angle is a rotation of the second and third principal directions about the axis of the first principal direction. For this two-dimensional, cross-sectional problem, the first and third angles are set to zero because they represent three-dimensional effects, and the second angle is set equal to the dip angle ($30^{\circ}$).

Layered-connectivity grids produce results that deviate substantially from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), whereas full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved substantial error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

@@

Our idealized benchmark problem serves as a reminder of the assumptions and limitations inherent in the use of vertically offset grids and points to conditions under which full cell connectivity might be of benefit in practical modeling studies. When such conditions exist in a MODFLOW 6 model, investigation of the effect of cell connectivity may be warranted. In a complex, real-world model, uncertainty in the structure and properties of hydrogeologic units, stresses on the system, and other model parameters could lead to compensation for the effects of inadequate cell connectivity during model calibration. However, when unrealistic flows or parameter values are observed in a calibrated model, inadequate cell connectivity can be considered among the possible sources of structural error, and full connectivity can be considered as a remedy.

\section*{Acknowledgments}
Kerry Bardot is supported by The Australian Research Council, The Western Australian Department of Water and Environmental Regulation and The Water Corporation of Western Australia through grant number LP18101153. The authors sincerely thank James Rumbaugh, John Masterson, Kalle Jahn, Jacob LaFontaine, John Brakebill, Heather Welch, Randall Hunt, and Richard Yager for their review comments, which improved the clarity of the manuscript and refined our conclusions regarding the practical implications of this study. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

>>>>

In simulations of groundwater flow through dipping aquifers, layers of model cells are often ``deformed'' to follow the top and bottom elevations of the aquifers. When this approach is used in MODFLOW, adjacent cells within the same model layer are vertically offset from one another, and the standard conductance-based (two-point) formulation for flow between cells does not rigorously account for these offsets. The XT3D multi-point flow formulation introduced in MODFLOW 6 is designed to account for geometric irregularities in the grid, including vertical offsets, and to provide accurate results for both isotropic and anisotropic groundwater flow.  A recent study evaluated the performance of the standard formulation and XT3D using a simple, synthetic benchmark model of a steeply dipping aquifer. Although XT3D generally improved the accuracy of flow simulations relative to the standard formulation as expected, neither formulation produced accurate flows in cases that involved large vertical offsets. In this paper, we explain that the inability of XT3D to produce accurate flows in the steeply dipping aquifer benchmark was not due to an inherent limitation of the flow formulation, but rather to the limited cell connectivity inherent in the most commonly used discretization packages in MODFLOW 6. Furthermore, we demonstrate that XT3D is able to produce the expected accuracy when adequate cell connectivity is introduced using MODFLOW's unstructured grid type and the aquifer is discretized vertically using at least two model layers.

@@

Numerical simulation of layered aquifer systems can involve a tradeoff between efficiency and accuracy. A traditional approach to discretizing layered aquifer systems, called the ``grid overlay'' method by \cite{hoaglund2003}, maps the heterogeneous hydrologic properties onto a rectilinear grid \citep{modflow84}. While this method preserves the benefits of the rectilinear grid structure, it can require a large number of model grid layers to adequately discretize hydrogeogic layers of variable dip and thickness \citep{Zyvoloski2006}.

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This commonly used method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. However, the non-orthogonality of the resulting grid can be a source of significant model error \citep{hoaglund2003}.

@@

The introduction of unstructured grids into MODFLOW \citep{modflowusg, modflow6gwf} presented new ways for a grid to violate the CVFD requirements. To help overcome the limitations of the two-point flow formulation on geometrically irregular grids, MODFLOW-USG and MODFLOW 6 offer a Ghost-Node Correction (GNC) Package \citep{modflowusg, modflow6gwf}. However, configuration of optimal ghost node locations and interpolation weights is highly problem-dependent, and we are not aware of guidelines for handling general unstructured grids or arbitrarily oriented 3D anisotropy.

The optional XT3D flow formulation \citep{modflow6xt3d} introduced in MODFLOW 6 offers a more ``automatic" alternative to ghost nodes. XT3D is a multi-point flow formulation \citep{edwards1998, aavatsmark2002} based on gradient reconstruction \citep{mavriplis2003leastsq, diskin2008accuracy}. By performing flow calculations using Darcy's Law in its tensorial form using a multi-point approximation of the head gradient vector, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid. In theory, XT3D should be able to compensate for the geometric irregularity introduced by vertical offsets between adjacent cells in the same grid layer.

@@

Versions of MODFLOW prior to MODFLOW-USG \citep{modflowusg} supported only structured grids composed of rows, columns, and layers of cells, which could be vertically offset. This grid type, called ``DIS," continues to be supported in MODFLOW-USG and MODFLOW 6, which also offer an unstructured grid type called ``DISU" \citep{modflowusg, modflow6gwf}. In addition, MODFLOW 6 offers an  unstructured grid type called ``DISV," which is based on specification of cell vertices \citep{modflow6gwf}. The DIS and DISV grid types are based on model layers, with the same horizontal grid structure within each layer. In a DIS or DISV grid, a connection between two cells in different model layers is created automatically if and only if one cell is directly below the other and neither cell has been optionally excluded from the model domain by the user. The DISU grid type is more flexible and, in MODFLOW 6, does not require the grid to have a layered structure. (The DISU grid type in MODFLOW-USG is similarly flexible but retains the concept of layers.) In a DISU grid, the user explicitly specifies all connections between grid cells. Thus, when used to represent a grid with a layered structure, as in this work, the DISU grid type allows connections to exist between any two cells, whether they are in the same layer or different layers. Grid geometry and connectivity information supplied by the discretization package (DIS, DISV, or DISU) is used by the MODFLOW 6 Node Property Flow (NPF) Package to formulate expressions for groundwater flow between cells in terms of heads calculated at cells centers. By default, the NPF Package uses a two-point flow formulation, which we call the ``conductance-based'' formulation. The optional, multi-point XT3D flow formulation can be activated simply by including the corresponding keyword in the NPF Package input.

@@

Note how the vertically offset grid violates the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same grid layer are not strictly horizontal and therefore not perpendicular to the vertical cell-cell interfaces they intersect. For example, in Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell-cell interface, which reduces the accuracy of the conductance-based formulation.

@@

Specific discharge at cell centers is calculated in a postprocessing step using distance-weighted interpolation of face flows calculated by MODFLOW 6 and reported in the binary budget output file. For the purpose of illustrating and evaluating the results of this idealized benchmark, specific discharge in cells within the aquifer is calculated using only face flows between cells within the aquifer. Likewise, specific discharge in cells within the surrounding domain is calculated using only face flows between cells within the domain. These specific discharge values are different from those reported in the binary budget output file by MODFLOW 6 because MODFLOW interpolates specific discharge in a cell using all of the face flows for that cell, including flows across cell faces on the aquifer-domain boundary. The flows reported by MODFLOW represent an ``average'' of the different flows on either side of the aquifer-domain boundary. Therefore, specific discharge calculated in a cell adjacent to the aquifer-domain boundary is affected by the different flow rate on the other side of the boundary. While this is an unavoidable consequence of vertically offset discretization in practical applications, and not necessarily a problem, users should be aware of these effects when interpreting the results of MODFLOW 6 dispersive-transport and MODPATH particle-tracking simulations.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows significantly improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c), which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively. Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d), reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.

@@

The results in Figure \ref{fig:fig3} confirm that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively. When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no significant benefit in our benchmark problem (Figure \ref{fig:fig3}b).

@@

The base case and variations on the base case up to this point have used isotropic hydraulic conductivity. However, sedimentary layers often exhibit anisotropic conductivity. Therefore, we examine the effect of anisotropy within the aquifer and domain by dividing the original conductivity perpendicular to the dip of the aquifer by factors of 1, 10, 100, 1000 and 10,000 to give anisotropy ratios ranging from 1.0 (isotropic) to 0.0001. Orientation of the principal directions of anisotropy parallel and perpendicular to the dip of the aquifer is accomplished by setting three angles in the NPF Package input. The first angle is the rotation within the horizontal plane of the first principal direction (typically the direction of maximum hydraulic conductivity) with respect to the positive $x$ axis. The second angle is the rotation of the first principal direction upward from the horizontal plane. The third angle is a rotation of the second and third principal directions about the axis of the first principal direction. For this two-dimensional, cross-sectional problem, the first and third angles are set to zero because they represent three-dimensional effects, and the second angle is set equal to the dip angle ($30^{\circ}$).

Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate significantly from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved significant error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

@@

Our idealized benchmark problem serves as a reminder of the assumptions and limitations inherent in the use of vertically offset grids and points to conditions under which full cell connectivity might be of benefit in practical modeling studies. When such conditions exist in a MODFLOW 6 model, investigation of the effect of cell connectivity may be warranted. In a complex, real-world model, uncertainty in the structure and properties of hydrogeologic units, stresses on the system, and other model parameters could lead to compensation for the effects of inadequate cell connectivity during model calibration. However, when unrealistic flows or parameters values are observed in a calibrated model, inadequate cell connectivity can be considered among the possible sources of structural error, and full connectivity can be considered as remedy.

\section*{Acknowledgments}
Kerry Bardot is supported by The Australian Research Council, The Western Australian Department of Water and Environmental Regulation and The Water Corporation of Western Australia through grant number LP18101153. The authors sincerely thank James Rumbaugh, John Masterson, Kalle Jahn, Jacob LaFontaine, Randall Hunt, and Richard Yager for their review comments, which improved the clarity of the manuscript and refined our conclusions regarding the practical implications of this study.

### 2024-11-19 latex/saved drafts/article_clean_v2.tex
intermediate review drafts & final draft in progress
intermediate review drafts & final draft in progress

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This commonly used method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. However, the non-orthogonality of the resulting grid can be a source of significant model error \citep{hoaglund2003}.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows significantly improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c), which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively. Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d), reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.

@@

The results in Figure \ref{fig:fig3} confirm that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively. When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no significant benefit in our benchmark problem (Figure \ref{fig:fig3}b).

@@

Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate significantly from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved significant error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

>>>>

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This commonly used method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. However, the non-orthogonality of the resulting grid can be a source of substantial model error \citep{hoaglund2003}.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows substantially improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c), which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively. Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d), reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.

@@

The results in Figure \ref{fig:fig3} confirm that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively. When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no substantial benefit in our benchmark problem (Figure \ref{fig:fig3}b).

@@

Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate substantially from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved substantial error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

### 2024-11-19 latex/saved drafts/article_tracked_changes.tex
intermediate review drafts & final draft in progress
intermediate review drafts & final draft in progress

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This {\color{red} commonly used} method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. {\color{red} \sout{In fact, it is common practice to represent each hydrogeologic layer using a single, boundary-matching model grid layer. }}However, the non-orthogonality of the resulting grid can be a source of {\color{red} \sout{significant}substantial} model error \citep{hoaglund2003}.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows {\color{red} \sout{significantly}substantially} improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c){\color{red}, which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively.} Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d){\color{red} \sout{.}, reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.}

@@

{\color{red} \sout{Like the results in Figure \ref{fig:fig2}, t}T}he results in Figure \ref{fig:fig3} {\color{red} \sout{show}confirm} that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. {\color{red}For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively.} When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. {\color{red} With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no substantial benefit in our benchmark problem (Figure \ref{fig:fig3}b).}

@@

Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate {\color{red} \sout{significantly}substantially} from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). {\color{red} For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively.} {\color{red} These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.}

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved {\color{red} \sout{significant}substantial} error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

>>>>

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This {\color{red} commonly used} method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. {\color{red} \sout{In fact, it is common practice to represent each hydrogeologic layer using a single, boundary-matching model grid layer. }}However, the non-orthogonality of the resulting grid can be a source of significant model error \citep{hoaglund2003}.

@@

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows significantly improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c){\color{red}, which overestimates the flux magnitude and total flow by approximately 3\% and 9\%, respectively.} Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d){\color{red} \sout{.}, reducing the errors in flux magnitude and total flow to approximately $-2 \times 10^{-4}$\% and $2 \times 10^{-4}$\%, respectively.}

@@

{\color{red} \sout{Like the results in Figure \ref{fig:fig2}, t}T}he results in Figure \ref{fig:fig3} {\color{red} \sout{show}confirm} that the use of full connectivity and the XT3D flow formulation reduces the errors in calculated flows, provided at least two grid layers are used to vertically discretize the aquifer. {\color{red}For example, when the aquifer is represented using two layers with layered connectivity, the flow error is approximately 18\% and 5\% for the standard and XT3D flow formulations, respectively. Use of full connectivity reduces these errors to approximately 11\% and $3 \times 10^{-4}$\%, respectively.} When a single grid layer is used to represent the aquifer, the enhanced connectivity needed to achieve an accurate solution is not possible, and the XT3D flow formulation is unable to compensate for the inadequate connectivity. {\color{red} With full connectivity and the XT3D flow formulation, using more than two layers to discretize the aquifer provides no significant benefit in our benchmark problem (Figure \ref{fig:fig3}b).}

@@

Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate significantly from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). {\color{red} For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively.} {\color{red} These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.}

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved significant error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.
